Foundations of a live data exploration environment

### 2020-03-09 programming-2020/paper.tex
Move published paper

\documentclass[english,crc]{programming}
\usepackage[backend=biber]{biblatex}
\addbibresource{paper.bib}
\usepackage{multicol}
\usepackage{microtype}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumitem}

\newcounter{thc}
\newcounter{dfc}
\theoremstyle{plain}
\newtheorem{lem}[thc]{Lemma}
\newtheorem{theorem}[thc]{Theorem}

\theoremstyle{definition}
\newtheorem{axiom}[dfc]{Axiom}
\newtheorem{definition}[dfc]{Definition}

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{idclr}{rgb}{0.0,0.0,0.0}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

@@

\begin{document}
\title{Foundations of a live data exploration environment}
%\titlerunning{Learning to live with errors}
\author[a,b]{Tomas Petricek}
\authorinfo{
\noindent
Tomas is a lecturer at University of Kent. He is working on making programming and data science
easier and more accessible. His research interests span a range of areas including theory of
programming languages, tools for data science as well as history and philosophy of programming.
%
Previously, he contributed to functional programming and the development of the F\# language and
type providers at Microsoft Research and obtained PhD from University of Cambridge for his work
on coeffects, a theory of context-aware programming languages. Along the way, he became interested
in understanding programming through the perspective of history and philosophy of science and wrote
papers about the evolution of programming concepts such as types and errors.
%
Contact him at \email{tomas@tomasp.net}.}
\affiliation[a]{University of Kent, UK}
\affiliation[b]{and The Alan Turing Institute, UK}
\keywords{Data exploration, live programming, data journalism, instant previews}
\paperdetails{
  perspective=sciencetheoretical,
  area={Programming environments},
}
% https://dl.acm.org/ccs/ccs.cfm and generate your Classification
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002951.10003227.10003351</concept_id>
<concept_desc>Information systems~Data mining</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011041</concept_id>
<concept_desc>Software and its engineering~Compilers</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10003120.10003121.10003129</concept_id>
<concept_desc>Human-centered computing~Interactive systems and tools</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[300]{Human-centered computing~Interactive systems and tools}
\ccsdesc[300]{Information systems~Data mining}
\ccsdesc[300]{Software and its engineering~Compilers}
\maketitle

@@

An increasing number of programming environments aim to provide a live development experience
for standard programming languages, but doing this is not easy. Fully recomputing the whole program
after every keystroke is inefficient and calculating how a change in the source code changes the
result is extremely hard when the text editor allows arbitrary changes.
Consider the following snippet that gets the release years of the $10$ most expensive movies from a
data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} =\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
\end{array}
\end{equation*}
%
A live programming environment computes and displays the list of years. Suppose that the programmer then
makes $\num{10}$ a variable and changes the date format:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\ident{count}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))\\
\end{array}
\end{equation*}
%
Ideally, the live programming environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking the first 10 elements) and only evaluate the
\ident{map} operation to differently format the release dates of top 10 movies. Our environment
does this for a simple data exploration language in an unrestricted text editor. We discuss related
work in Section~\ref{sec:future}, but we briefly review the most important directions here, in
order to situate our contributions.

@@

\begin{itemize}
\item We introduce the \emph{data exploration calculus} (Section~\ref{sec:calculus}), a small formally
  tractable language for data exploration. The calculus is motivated by our review
  of how data analysts work (Section~\ref{sec:background}) and makes it possible to write simple,
  transparent and reproducible data analyses.

\item A live programming environment does not operate in batch mode and so we cannot follow classic
  compiler literature. We capture the essence of such new perspective (Section~\ref{sec:formal})
  and use it to build our \emph{instant preview} mechanism (Section~\ref{sec:previews}) that evaluates
  code instantly during editing.

\item We prove that our instant preview mechanism is correct (Section~\ref{sec:evaluation-correctness}) and that
  it reuses previously evaluated values when possible (Section~\ref{sec:evaluation-reuse}). We
  illustrate the practicality of the mechanism using an empirical evaluation
  (Section~\ref{sec:evaluation-empirical}) and a case study (Section~\ref{sec:evaluation-case}).
\end{itemize}

@@

\begin{figure}[b]
\hspace{1em}\includegraphics[scale=0.52]{figures/notebook.png}
\caption{Financial Times analysis that joins UN trade database with ISO country codes.}
\label{fig:ft-uncomtrade}
\end{figure}

@@

Data scientists often use general-purpose programming languages such as Python, but the kind of
code they write and the way they interact with the system is very different from how software
engineers work \cite{workflow}. This paper focuses on simple data wrangling and data exploration as
done, for example, by journalists analysing government datasets. This new kind of non-expert
programmers is worth our attention as they often work on informing the public. They need easy-to-use
tools, but not necessarily a full programming langugae. In this section, we ilustrate how such
data analyses look and we provide a justification for the design of our data exploration calculus.

@@

Consider the Financial Times analysis of plastic waste \cite{ftnotebooks,ftarticle}. It joins
datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
an excerpt from one notebook of the data analysis. The code has a number of important properties:
%
\begin{itemize}
\vspace{0.5em}
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any custom functions. Code is parameterized by having a global variable
  \ident{material} set to \str{plastics} and keeping other possible values in a comment.
  This lets the analyst run and check results of intermediate steps.

\vspace{0.5em}
\item The code relies on external libraries. Our example uses Pandas \cite{pandas},
  which provides operations for data wrangling such as \ident{merge} to join datasets
  or \ident{drop\_duplicates} to delete rows with duplicate column values. Such standard libraries
  are external to the data analysis and are often implemented in another language like C++.

\vspace{0.5em}
\item The code is structured as a sequence of commands. Some commands define a variable, either by
  loading data, or by transforming data loaded previously. Even in Python, data is often treated
  as immutable. Other commands produce an output that is displayed in the notebook.

\vspace{0.5em}
\item There are many corner cases, such as the fact that \ident{keep\_default\_na}
  needs to be set to handle Namibia correctly. These are discovered interactively by
  running the code and examining the output, so providing an instant feedback is essential.
\end{itemize}

\vspace{0.5em}
\noindent
Many Jupyter notebooks are more complex than the above example and might use helper
functions or object-oriented code. However, simple data analyses such as the one discussed
here are frequent enough and pose interesting problems for programming tools.
This paper aims to bring such analyses to the attention of programming research community
by capturing their essential properties as a formal calculus.

@@

\begin{figure}
\centering
\subcaptionbox{
  \raggedright\footnotesize The analysis counts the number of distinct events per
    athlete. After typing `.' the editor offers further aggregation operations.
  \label{fig:gamma-dot}
}{\includegraphics[scale=0.26]{figures/gamma1.png}}
\subcaptionbox{
  \raggedright\footnotesize
    Our live programming environment for The Gamma. The table is updated on-the-fly a
    and shows the result at the current cursor position.
  \label{fig:gamma-preview}
}{\includegraphics[scale=0.26]{figures/gamma2.png}}
\vspace{0.5em}
\caption{\small Previous work on The Gamma with auto-complete based on type information (left)
  and our new editor with instant preview (right).}
\label{fig:gamma-intro}
\end{figure}

@@

Simple data exploration has been the motivation for
a scripting language The Gamma \cite{gamma}. Scripts in The Gamma are sequences of commands
that either define a variable or produce an output. It does not support top-level
functions and lambda functions can be used only as method arguments.
Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
such as JavaScript. The Gamma uses type providers \cite{providers-fsharp}
for accessing external data sources.
Type providers generate object types with members and The Gamma offers those in an
auto-complete list when the user types dot (`.') to access a member.
The combination of type providers and auto-complete makes it
possible to solve a large number of data exploration tasks through the very simple interaction of
selecting operations from a list.

The example in Figure~\ref{fig:gamma-dot} summarizes data on Olympic medals. Identifiers such
as \qident{sum Bronze} are names of members generated by the type provider. The type provider
used in this example generates an object with members for data transformations such as
\qident{group data}, which return further objects with members for specifying transformation
parameters, such as selecting the grouping key using \qident{by Athlete}.

The Gamma language is richer, but the example in Figure~\ref{fig:gamma-dot} shows that non-trivial
data exploration can be done using a very simple language. The assumptions about structure of
code that are explicit in The Gamma are implicitly present in Python and R data analyses
produced by journalists, economists and other users with other than programming background.
When we refer to The Gamma in this paper, readers not familiar with it can consider a small
subset of Python.

@@

The implementation that accompanies this paper builds a live programming environment for The Gamma.
It is discussed in Section~\ref{sec:evaluation-case} and it replaces
the original text editor with just auto-complete with a live programming environment
that provides \emph{instant previews}.

An example of a instant preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
Gamma programs consist of lists of commands which are either expressions or let bindings. Our
editor displays a instant preview below the command that the user is currently editing. The preview
shows the result of evaluating the expression or the value assigned to a bound variable. When the
user changes the code, the preview is updated automatically, without any additional interaction
with the user.

@@

The \emph{data exploration calculus} is a small formal language for data exploration. The calculus
is not, in itself, Turing-complete and it can only be used together with external libraries that
define what objects are available and what the behaviour of their members is. This is
sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
define the calculus in this section and then use it to formalise our instant preview
mechanism in Section~\ref{sec:formal}.
%
The instant preview mechanism does not rely on types and so we postpone the discussion of static
typing to Appendix~\ref{sec:extra-types}. Interestingly, it reuses the mechanism used for live
previews.

@@

\begin{figure}
\raggedright
\hspace{0.5em}{\small\sffamily Programs, commands, terms, expressions and values}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{lcl}
p &\narrow{::=}& c_1; \ldots; c_n\\
c &\narrow{::=}& \kvd{let}~x = t \lsep t\\
\end{array}
\qquad
\begin{array}{lcl}
t &\narrow{::=}& o \lsep x \\
  &\narrow{|}& t.m(e, \ldots, e)
\end{array}
\qquad
\begin{array}{lcl}
e &\narrow{::=}& t \lsep \lambda x\rightarrow e\\
v &\narrow{::=}& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

\vspace{0.5em}
\hspace{0.5em}{\small\sffamily Evaluation contexts of expressions}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{lcl}
C_e[-] &=& C_e[-].m(e_1, \ldots, e_n) ~\lsep~ o.m(v_1, \ldots, v_{m}, C_e[-], e_1,\ldots , e_n) ~\lsep~ -\\
C_c[-] &=& \kvd{let}~x = C_e[-] ~\lsep~ C_e[-]\\
C_p[-] &=& o_1;\; \ldots;\; o_{k};\; C_c[-];\; c_{1};\; \ldots;\; c_n\\
\end{array}
\end{equation*}

\vspace{0.5em}
\hspace{0.5em}{\small\sffamily Let elimination and member reduction}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{ll}
\hspace{-0.5em}\begin{array}{l}
o_1;\; \ldots;\; o_{k};\; \kvd{let}~x=o;\; c_{1};\; \ldots; c_n \rightsquigarrow\\
\qquad  o_1;\; \ldots;\; o_{k};\; o; c_{1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
\quad\text{\rname{let}}\\[1.2em]
\inference
  {o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}
  {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} &
\quad\text{\rname{external}}
\end{array}
\end{equation*}

@@

\subsection{Language syntax}
The calculus combines object-oriented features such as member access with functional features
including lambda functions. The syntax is defined in Figure~\ref{fig:dec-calculus}. Object values
$o$ are defined by external libraries that are used in conjunction with the core calculus.

@@

The data exploration calculus is a call-by-value language. We model evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ (one object for each command, including let bindings) which can be
displayed as intermediate results of the data analysis. The operational semantics is parameterized
by a relation $\rightsquigarrow_\epsilon$ that models the functionality of the external libraries
used with the calculus and defines the reduction behaviour for member accesses. The relation has the
following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
Here, the operation $m$ is invoked on an object and takes values (objects or function values) as arguments.
The reduction always results in an object.
Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evaluation of a command and $C_p$ defines left-to-right evaluation of a program.
The rule \rname{external} calls a method provided by an external library in a
call-by-value fashion; \rname{let} substitutes a value of an evaluated variable in
all subsequent commands and leaves the result in the list of commands.
Note that our semantics does not define how $\lambda$ applications are reduced. This is done by
external libraries, which will typically supply functions with arguments using standard $\beta$-reduction.
The behaviour is subject to constraints discussed next.

@@

The key idea behind our method is to maintain a dependency graph \cite{dependencies} with
nodes representing individual operations of the computation that can be evaluated
to obtain a preview. Each time the program text is modified, we parse it afresh (using an
error-recovering parser) and bind the abstract syntax tree to the dependency graph.
When binding a new expression to the graph, we reuse previously created nodes as long as
they have the same structure and the same dependencies. For expressions that have a new
structure, we create new nodes.

@@

\subsubsection{Example graph}
Figure~\ref{fig:dep-graph} illustrates how we build the
dependency graph. Node representing $\ident{take}(x)$ depends on the argument -- the
number $\num{15}$ -- and the instance, which is a node representing $\ident{skip}(\num{10})$.
This depends on the instance \ident{data} and the number $\num{10}$. Note that variables
bound via \kvd{let} binding such as $x$ do not appear as $\bnd{var}$ nodes. The node using it
depends directly on the node representing the expression assigned to $x$.

@@

In this section, we describe how previews are evaluated. The evaluation is done over the dependency
graph, rather than over the structure of program expressions as in the operational semantics
given in Section~\ref{sec:calculus-semantics}. In Section~\ref{sec:evaluation}, we prove that
resulting previews are the same as the result we would get by directly evaluating code and we
also show that no recomputation occurs when code is edited in certain ways.

@@

\newpage
\appendix

>>>>

\AtBeginDocument{\paperdetails{
    submitted=2019-09-30,
    published=2020-02-17,
    year=2020,
    volume=4,
    issue=3,
    articlenumber=8,
  }}
\documentclass[english,crc,references=cleveref]{programming}
\usepackage[backend=biber]{biblatex}
\addbibresource{paper.bib}
% \usepackage{multicol}
% \usepackage{microtype}
% \usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
% \usepackage{amsmath}
\usepackage{amsthm}
% \usepackage{enumitem}

\AtBeginDocument{
  \newcounter{thc}
\newcounter{dfc}
\theoremstyle{plain}
\newtheorem{lem}[thc]{Lemma}
\newtheorem{theorem}[thc]{Theorem}

\theoremstyle{definition}
\newtheorem{axiom}[dfc]{Axiom}
\newtheorem{definition}[dfc]{Definition}
}

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{idclr}{rgb}{0.0,0.0,0.0}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{\;\;|\;\;}
% \newcommand{\Num}[1]{\textcolor{numclr}{#1}}
\newcommand{\Num}[1]{\num[color=numclr]{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

@@

\begin{document}
\title{Foundations of a live data exploration environment}
%\titlerunning{Learning to live with errors}
\author[a,b]{Tomas Petricek}
\authorinfo{%
%
 % Tomas
  is a lecturer at University of Kent. He is working on making programming and data science
easier and more accessible. His research interests span a range of areas including theory of
programming languages, tools for data science as well as history and philosophy of programming.
%
Previously, he contributed to functional programming and the development of the F\# language and
type providers at Microsoft Research and obtained PhD from University of Cambridge for his work
on coeffects, a theory of context-aware programming languages. Along the way, he became interested
in understanding programming through the perspective of history and philosophy of science and wrote
papers about the evolution of programming concepts such as types and errors.
%
Contact him at \email{tomas@tomasp.net}.}
\affiliation[a]{University of Kent, UK}
\affiliation[b]{and The Alan Turing Institute, UK}
\keywords{Data exploration, live programming, data journalism, instant previews}
\paperdetails{
  perspective=sciencetheoretical,
  area={Programming environments},
}
% https://dl.acm.org/ccs/ccs.cfm and generate your Classification
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002951.10003227.10003351</concept_id>
<concept_desc>Information systems~Data mining</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011041</concept_id>
<concept_desc>Software and its engineering~Compilers</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10003120.10003121.10003129</concept_id>
<concept_desc>Human-centered computing~Interactive systems and tools</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[300]{Human-centered computing~Interactive systems and tools}
\ccsdesc[300]{Information systems~Data mining}
\ccsdesc[300]{Software and its engineering~Compilers}
\maketitle

@@

An increasing number of programming environments aim to provide a live development experience
for standard programming languages, but doing this is not easy. Fully recomputing the whole program
after every keystroke is inefficient and calculating how a change in the source code changes the
result is extremely hard when the text editor allows arbitrary changes.
Consider the following snippet that gets the release years of the $10$ most expensive movies from a
data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} =\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\Num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
\end{array}
\end{equation*}
%
A live programming environment computes and displays the list of years. Suppose that the programmer then
makes $\Num{10}$ a variable and changes the date format:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{count} = \Num{10}\\
\kvd{let}~\ident{top} = \ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\ident{count}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))\\
\end{array}
\end{equation*}
%
Ideally, the live programming environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking the first 10 elements) and only evaluate the
\ident{map} operation to differently format the release dates of top 10 movies. Our environment
does this for a simple data exploration language in an unrestricted text editor. We discuss related
work in \cref{sec:future}, but we briefly review the most important directions here, in
order to situate our contributions.

@@

\begin{itemize}
\item We introduce the \emph{data exploration calculus} (\cref{sec:calculus}), a small formally
  tractable language for data exploration. The calculus is motivated by our review
  of how data analysts work (\cref{sec:background}) and makes it possible to write simple,
  transparent and reproducible data analyses.

\item A live programming environment does not operate in batch mode and so we cannot follow classic
  compiler literature. We capture the essence of such new perspective (\cref{sec:formal})
  and use it to build our \emph{instant preview} mechanism (\cref{sec:previews}) that evaluates
  code instantly during editing.

\item We prove that our instant preview mechanism is correct (\cref{sec:evaluation-correctness}) and that
  it reuses previously evaluated values when possible (\cref{sec:evaluation-reuse}). We
  illustrate the practicality of the mechanism using an empirical evaluation
  (\cref{sec:evaluation-empirical}) and a case study (\cref{sec:evaluation-case}).
\end{itemize}

@@

\begin{figure}
  \begin{wide}
    \includegraphics[width=\linewidth]{figures/notebook}
  \end{wide}
  \caption{Financial Times analysis that joins UN trade database with ISO country codes.}
\label{fig:ft-uncomtrade}
\end{figure}

@@

Data scientists often use general-purpose programming languages such as Python, but the kind of
code they write and the way they interact with the system is very different from how software
engineers work~\cite{workflow}. This paper focuses on simple data wrangling and data exploration as
done, for example, by journalists analysing government datasets. This new kind of non-expert
programmers is worth our attention as they often work on informing the public. They need easy-to-use
tools, but not necessarily a full programming langugae. In this section, we ilustrate how such
data analyses look and we provide a justification for the design of our data exploration calculus.

@@

Consider the Financial Times analysis of plastic waste~\cite{ftnotebooks,ftarticle}. It joins
datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. \Cref{fig:ft-uncomtrade} shows
an excerpt from one notebook of the data analysis. The code has a number of important properties:
%
\begin{itemize}
%\vspace{0.5em}
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any custom functions. Code is parameterized by having a global variable
  \ident{material} set to \str{plastics} and keeping other possible values in a comment.
  This lets the analyst run and check results of intermediate steps.

%\vspace{0.5em}
\item The code relies on external libraries. Our example uses Pandas~\cite{pandas},
  which provides operations for data wrangling such as \ident{merge} to join datasets
  or \ident{drop\_duplicates} to delete rows with duplicate column values. Such standard libraries
  are external to the data analysis and are often implemented in another language like C++.

%\vspace{0.5em}
\item The code is structured as a sequence of commands. Some commands define a variable, either by
  loading data, or by transforming data loaded previously. Even in Python, data is often treated
  as immutable. Other commands produce an output that is displayed in the notebook.

%\vspace{0.5em}
\item There are many corner cases, such as the fact that \ident{keep\_default\_na}
  needs to be set to handle Namibia correctly. These are discovered interactively by
  running the code and examining the output, so providing an instant feedback is essential.
\end{itemize}
%
%\vspace{0.5em}
%
Many Jupyter notebooks are more complex than the above example and might use helper
functions or object-oriented code. However, simple data analyses such as the one discussed
here are frequent enough and pose interesting problems for programming tools.
This paper aims to bring such analyses to the attention of programming research community
by capturing their essential properties as a formal calculus.

@@

\begin{figure}
  \begin{wide}\centering
    \begin{subfigure}{.49\linewidth}
      \includegraphics[width=\linewidth]{figures/gamma1.png}
      \caption{The analysis counts the number of distinct events per athlete.
        After typing `.' the editor offers further aggregation operations.}
  \label{fig:gamma-dot}
    \end{subfigure}\hfill
    \begin{subfigure}{.49\linewidth}
      \includegraphics[width=\linewidth]{figures/gamma2.png}
      \caption{Our live programming environment for The Gamma. The table is
        updated on-the-fly and shows the result at the current cursor
        position.}
  \label{fig:gamma-preview}
    \end{subfigure}
  \end{wide}
% \vspace{0.5em}
\caption{Previous work on The Gamma with auto-complete based on type information (left)
  and our new editor with instant preview (right).}
\label{fig:gamma-intro}
\end{figure}

@@

\begin{sloppypar}
Simple data exploration has been the motivation for
a scripting language The Gamma~\cite{gamma}. Scripts in The Gamma are sequences of commands
that either define a variable or produce an output. It does not support top-level
functions and lambda functions can be used only as method arguments.
Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
such as JavaScript. The Gamma uses type providers~\cite{providers-fsharp}
for accessing external data sources.
Type providers generate object types with members and The Gamma offers those in an
auto-complete list when the user types dot (`.') to access a member.
The combination of type providers and auto-complete makes it
possible to solve a large number of data exploration tasks through the very simple interaction of
selecting operations from a list.
\end{sloppypar}

The example in \cref{fig:gamma-dot} summarizes data on Olympic medals. Identifiers such
as \qident{sum Bronze} are names of members generated by the type provider. The type provider
used in this example generates an object with members for data transformations such as
\qident{group data}, which return further objects with members for specifying transformation
parameters, such as selecting the grouping key using \qident{by Athlete}.

The Gamma language is richer, but the example in \cref{fig:gamma-dot} shows that non-trivial
data exploration can be done using a very simple language. The assumptions about structure of
code that are explicit in The Gamma are implicitly present in Python and R data analyses
produced by journalists, economists and other users with other than programming background.
When we refer to The Gamma in this paper, readers not familiar with it can consider a small
subset of Python.

@@

The implementation that accompanies this paper builds a live programming environment for The Gamma.
It is discussed in \cref{sec:evaluation-case} and it replaces
the original text editor with just auto-complete with a live programming environment
that provides \emph{instant previews}.

An example of a instant preview is shown in \cref{fig:gamma-preview}. As noted earlier, The
Gamma programs consist of lists of commands which are either expressions or let bindings. Our
editor displays a instant preview below the command that the user is currently editing. The preview
shows the result of evaluating the expression or the value assigned to a bound variable. When the
user changes the code, the preview is updated automatically, without any additional interaction
with the user.

@@

The \emph{data exploration calculus} is a small formal language for data exploration. The calculus
is not, in itself, Turing-complete and it can only be used together with external libraries that
define what objects are available and what the behaviour of their members is. This is
sufficient to capture the simple data analyses discussed in \cref{sec:background}. We
define the calculus in this section and then use it to formalise our instant preview
mechanism in \cref{sec:formal}.
%
The instant preview mechanism does not rely on types and so we postpone the discussion of static
typing to \cref{sec:extra-types}. Interestingly, it reuses the mechanism used for live
previews.

@@

\begin{figure}
\raggedright
\hspace{0.5em}{\sffamily Programs, commands, terms, expressions and values}
%
\begin{equation*}
\begin{array}{lcl}
p &\narrow{::=}& c_1; \ldots; c_n\\
c &\narrow{::=}& \kvd{let}~x = t \lsep t\\
\end{array}
\qquad
\begin{array}{lcl}
t &\narrow{::=}& o \lsep x \\
  &\narrow{|}& t.m(e, \ldots, e)
\end{array}
\qquad
\begin{array}{lcl}
e &\narrow{::=}& t \lsep \lambda x\rightarrow e\\
v &\narrow{::=}& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

%
\hspace{0.5em}{\sffamily Evaluation contexts of expressions}
%
\begin{equation*}
\begin{array}{lcl}
C_e[-] &=& C_e[-].m(e_1, \ldots, e_n) ~\lsep~ o.m(v_1, \ldots, v_{m}, C_e[-], e_1,\ldots , e_n) ~\lsep~ -\\
C_c[-] &=& \kvd{let}~x = C_e[-] ~\lsep~ C_e[-]\\
C_p[-] &=& o_1;\; \ldots;\; o_{k};\; C_c[-];\; c_{1};\; \ldots;\; c_n\\
\end{array}
\end{equation*}

%
\hspace{0.5em}{\sffamily Let elimination and member reduction}
%
\begin{equation*}
\begin{array}{ll}
\hspace{-0.5em}\begin{array}{l}
o_1;\; \ldots;\; o_{k};\; \kvd{let}~x=o;\; c_{1};\; \ldots; c_n \rightsquigarrow\\
\qquad  o_1;\; \ldots;\; o_{k};\; o; c_{1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
\quad\text{\rname{let}}\\[1.2em]
\inference
  {o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}
  {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} &
\quad\text{\rname{external}}
\end{array}
\end{equation*}

@@

\subsection{Language syntax}
The calculus combines object-oriented features such as member access with functional features
including lambda functions. The syntax is defined in \cref{fig:dec-calculus}. Object values
$o$ are defined by external libraries that are used in conjunction with the core calculus.

@@

The data exploration calculus is a call-by-value language. We model evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ (one object for each command, including let bindings) which can be
displayed as intermediate results of the data analysis. The operational semantics is parameterized
by a relation $\rightsquigarrow_\epsilon$ that models the functionality of the external libraries
used with the calculus and defines the reduction behaviour for member accesses. The relation has the
following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
Here, the operation $m$ is invoked on an object and takes values (objects or function values) as arguments.
The reduction always results in an object.
\Cref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evaluation of a command and $C_p$ defines left-to-right evaluation of a program.
The rule \rname{external} calls a method provided by an external library in a
call-by-value fashion; \rname{let} substitutes a value of an evaluated variable in
all subsequent commands and leaves the result in the list of commands.
Note that our semantics does not define how $\lambda$ applications are reduced. This is done by
external libraries, which will typically supply functions with arguments using standard $\beta$-reduction.
The behaviour is subject to constraints discussed next.

@@

The key idea behind our method is to maintain a dependency graph~\cite{dependencies} with
nodes representing individual operations of the computation that can be evaluated
to obtain a preview. Each time the program text is modified, we parse it afresh (using an
error-recovering parser) and bind the abstract syntax tree to the dependency graph.
When binding a new expression to the graph, we reuse previously created nodes as long as
they have the same structure and the same dependencies. For expressions that have a new
structure, we create new nodes.

@@

\subsubsection{Example graph}
\cref{fig:dep-graph} illustrates how we build the
dependency graph. Node representing $\ident{take}(x)$ depends on the argument -- the
number $\Num{15}$ -- and the instance, which is a node representing $\ident{skip}(\Num{10})$.
This depends on the instance \ident{data} and the number $\Num{10}$. Note that variables
bound via \kvd{let} binding such as $x$ do not appear as $\bnd{var}$ nodes. The node using it
depends directly on the node representing the expression assigned to $x$.

@@

In this section, we describe how previews are evaluated. The evaluation is done over the dependency
graph, rather than over the structure of program expressions as in the operational semantics
given in \cref{sec:calculus-semantics}. In \cref{sec:evaluation}, we prove that
resulting previews are the same as the result we would get by directly evaluating code and we
also show that no recomputation occurs when code is edited in certain ways.

@@

\appendix
